---
title: "Geospatial Data & Maps in R, pt 2"
author: "Patty Frontiera, UC Berkeley D-Lab"
date: "May 2018"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part II Prep

1. https://github.com/dlab-geo/r-geospatial-workshop

- Click *Clone or Download* and download the zip file
- Upzip the zip file and make a note of the folder in which it is located

2. Open RStudio and start a **new script** or continue the one from last week

3. Set your working directory to where you downloaded and unzipped the files

4. Follow along by opening `r-geospatial-workshop-pt2.html` in a web browser

5. Make sure required libraries are installed. 

- sp, rgdal, rgeos, tmap, ggplot2, RColorbrewer, classInt

## Follow Along

Open one of the tutorial files in a web browser

Slides
`r-geospatial-workshop-pt2.html `

Tutorial
`r-geospatial-workshop-pt2-tutorial.html `

Raw Code
`scripts/r-geospatial-workshop-pt2-May-2018.R`

*Make sure you can cut and paste into RStudio*

## Part II Overview

Recap Part I

Map spatial objects with `tmap`

Taste of Spatial Analysis

## Part I Recap

In Part I, we:

- Loaded geospatial data from CSV files
- Mapped geospatial data with `ggplot` and `ggmap`
- Geocoded addresses with `ggmap`
- Promoted data frames to `SpatialPointsDataFrames` with `sp::coordinates`
- Loaded geodata from shapefiles with `rgdal::readOGR`
- Defined `CRSs` with `sp::proj4string` and `sp::CRS`
- Transformed CRSs with `sp::spTransform`
- Saved the sp object `sfboundary_lonlat` as a shapefile using `rgdal::writeOGR`

## R Spatial Libraries

Let's load the libraries we will use

```{r, eval=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read/write from file; manage CRSs
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
```

##

```{r, echo=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read and write from file
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
```
 
## Set your working directory

Use `setwd` to set your working directory to the location of the tutorial files.

For example:

```{r, eval=FALSE}
setwd("~/Documents/Dlab/workshops/2018/rgeo/r-geospatial-workshop/r-geospatial-workshop")
```


# Load files from Part I

## SF Properties 2015

```{r}
# Read in from CSV file
sfhomes <- read.csv('data/sf_properties_25ksample.csv', 
                    stringsAsFactors = FALSE)
# subset the data
sfhomes15 <- subset(sfhomes, as.numeric(SalesYear) == 2015)

sfhomes15_sp <- sfhomes15  # Make a copy

# Make it spatial
coordinates(sfhomes15_sp) <- c('lon','lat')

```

## BART Stations

```{r}
# Read in the Bart data from CSV file
bart <- read.csv("./data/bart.csv", stringsAsFactors = F)

```

## SF Boundary

Read in the `sf_boundary.shp` file with `rGDAL`

```{r}
sfboundary <- readOGR(dsn="data", layer="sf_boundary")

```

## Define & Transform the CRS

*Also called `defining` and `reprojecting`
```{r}
# Define the CRS
proj4string(sfhomes15_sp) <- CRS("+init=epsg:4326")

# Transform CRS
sfboundary_lonlat <- spTransform(sfboundary, CRS("+init=epsg:4326"))
```


## SF Highways

```{r}
# Read in the `sf_highways.shp` file with `rGDAL`
highways <- readOGR(dsn="data", layer="sf_highways")

# Transform to geographic coords
highways_lonlat <- spTransform(highways, CRS("+init=epsg:4326"))
```

# PART II 


# Mapping Spatial Objects

## So far we have created maps with

`base::plot`, `ggplot`, `ggmap` for geospatial data in data frames

- great for creating maps given these types of data


`sp::plot` for `sp` objects

- meh, but great for a quick look at spatial data

## There is also sp::spplot

Use it to create quick maps

Can be used to create great maps 

- BUT complex, non-intuitive syntax = long ugly code

See examples in [sp Gallery: Plotting maps with sp](https://edzer.github.io/sp)


# tmap

## tmap

`tmap` stands for thematic map

It's a powerful toolkit for creating maps with `sp` objects

yet, with less code than the alternatives

Syntax should be familar to ggplot2 users, but simpler

Relatively easy to create & save interactive maps


## tmap

Load the library

```{r}
library(tmap)
```
You may need to install /load dependencies
- ggplot2, RColorbrewer, classInt, leaflet libraries


## Quick tmap (qtm)

```{r}
qtm(sfboundary)
```

## Quick tmap (qtm)

Similar to the plot()

Now, make a `qtm` of `sfhomes15_sp`

```{r, eval=F}
qtm(sfboundary)
```


## tmap modes

`tmap` has 2 modes:

* `plot` for static maps

* `view` for interactive maps

Use the command `tmap_mode` to toggle between them:

* `tmap_mode("plot")`

* `tmap_mode("view")`

## Quick Interactive tmap

Create an interactive `tmap` by setting the mode to view.

Then recreate the `qtm` of `sfhomes15_sp`

```{r, eval=F}
tmap_mode("view") # set tmap to interactive view mode
qtm(sfhomes15_sp) # Interactive - click on the points
```

## Quick Interactive tmap

Check the `Popups` and `Layer selector`
```{r, echo=F}
tmap_mode("view") # set tmap to interactive view mode
qtm(sfhomes15_sp) # Interactive - click on the points
```

## Reset the mode to static plot

```{r}
tmap_mode("plot")
```


## Customizing tmaps

To customize your `tmap` you need to use the more complex syntax

```{r, eval=F}
tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black")
```

`tm_shape(<sp_object>)`  specifies the sp object

`+ tm_<element>(...)`  specifies the symbology

plus other options for creating a publication ready map

## Symbolizing polygons

```{r }
tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black")
```


## Mapping Points - `tm_dots`

`tm_dots` are a type of `tm_symbols()`

See `?tm_symbols` for other types of point symbols.

```{r, eval=FALSE}
tm_shape(sfhomes15_sp) + 
  tm_dots(col="red", size=.25)
```

## Mapping Points - `tm_dots`

```{r}
tm_shape(sfhomes15_sp) + 
  tm_dots(col="red", size=.25)
```

## Mapping Lines - `tm_lines`
```{r}
tm_shape(highways_lonlat) + 
  tm_lines(col="black")
```

## Mapping values

**Column names must be quoted!**
```{r}
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
```


## Mapping Multiple Layers

You can overlay multiple `sp` objects, or layers, on the map, by concatenating the `tmap` commands with plus signs

```{r, eval=F}
# Map the SF Boundary first
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +

# Overlay the highway lines next
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
  
# Then add the house points
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
 
```

## Mapping Multiple Layers

```{r, echo=F}
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
 
```

## Does this work

Try the code below 

Can you redo the previous map with `sfboundary` instead of  `sfboundary_lonlat`?

If yes, what does that tell you about `tmap`?

```{r, eval=F}
tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black") +
  
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
  
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25)
```

## 

```{r, echo=F}
tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25)
```

## CRS management in tmap

If the CRSs are defined, `tmap` will use that info - if not, `tmap` will assume `WGS84` -

and dynamically `reproject` subsequent layers to match first one added to map
```{r, echo=F}
tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black") +
  
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25) 
```

## Making it more Polished

```{r, eval=F}
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
  
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
  
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, 
          title = "San Francisco Property Values (2015)") + 
  
tm_layout(inner.margins=c(.05, .2, .15, .05)) 
      # bottom, left, top, right
```

## Making it more Polished

```{r, echo=F}
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") + 
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right
```


## Challenge

Make the previous `tmap` interactive

## Challenge - Solution
```{r, eval=F}
tmap_mode("view")

tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
  
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
  
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") + 
  
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right

# OR

#ttm()       # Toggle Tmap mode between "interactive" and "plot"
#last_map()  # display the last map 
```

## Challenge - Solution
```{r, echo=F}
tmap_mode("view")

tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") + 
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right

# OR

#ttm()       # Toggle Tmap mode between "interactive" and "plot"
#last_map()  # display the last map 
```

## Making it more pretty!
`tmap` also allows us to plot over a wide variety basemaps.

(Here's a list: http://leaflet-extras.github.io/leaflet-providers/preview/)

## Making it more pretty!
```{r}
tmap_mode("view")
watercolor_map <- tm_basemap("Stamen.Watercolor") +
tm_shape(sfhomes15_sp) + 
tm_dots(col="totvalue", size=.25, title = "San Francisco Property Values (2015)") +
tm_tiles("Stamen.TonerLabels")
```

## Making it more pretty!
```{r}
watercolor_map
```


## Customize the Popups

What changed in the code? Output map?

```{r, eval=F}
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
  
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +

tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, 
          title = "San Francisco Property Values (2015)",
          popup.vars=c("SalesYear","totvalue","NumBedrooms",
                       "NumBathrooms","AreaSquareFeet")) + 
  
tm_layout(inner.margins=c(.05, .2, .15, .05)) # bottom, left, top, right
```

## Customize the Popups

```{r, echo=F}
tm_shape(sfboundary_lonlat) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(highways_lonlat) + 
  tm_lines(col="black") +
tm_shape(sfhomes15_sp) + 
  tm_dots(col="totvalue", size=.25, 
          title = "San Francisco Property Values (2015)",
          popup.vars=c("SalesYear","totvalue","NumBedrooms",
                       "NumBathrooms","AreaSquareFeet"))
```


## Save the map

Save the map to a named variable (map1)
```{r }
map1 <- last_map()
map1 # then display it
```

## Save map to file

Use `save_tmap` and then take a look at your files

- The output file type depends on the file name extension!
```{r, eval=F}
save_tmap(map1, "sf_properties.png", height=6) # Static image file with
save_tmap(map1, "sf_properties.html") # interactive web map
```

```{r, echo=F}
save_tmap(map1, "sf_properties.png", height=6) # Static image file
save_tmap(map1, "sf_properties.html") # interactive web map
```

## tmap starting points

?tmap

?tmap_shape

?tmap_element

- ?tm_polygons (tm_fill, tm_borders)
- ?tm_symbols (tm_dots, etc...)

- [tmap in a Nutshell](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html) 

## Publish your Map

Many ways to do this.

You can share you map online by publishing it to [RPubs](https://rpubs.com).

- You need to have an `RPubs` account to make that work.

1. Enter the name of your tmap object (`map1`) or your  `tmap` code in the console

2. In the `Viewer` window, click on the **Publish** icon.


## RPub Demo

```{r, eval=F}
tmap_mode("view")
map1 #  in console
```

[http://rpubs.com/pfrontiera/358110](http://rpubs.com/pfrontiera/358110)

# Questions?

# Spatial Analysis

## Transform data to common CRS

In order to perform spatial analysis we need to first convert all data objects to a common CRS.

Which type? Projected or Geographic CRS?


## Geographic vs. Projected CRS

If my goal is to create maps, I may convert all data to a geographic CRS.

- Why?  Which one?

If my goal is to do spatial analysis, I will convert to a projected CRS.

- Why? Which one?


## Common CRS EPSG Codes

**Geographic CRSs**

* `4326` Geographic, WGS84 (default for lon/lat)

* `4269` Geographic, NAD83 (USA Fed agencies like Census)


**Projected CRSs**

* `5070` USA Contiguous Albers Equal Area Conic

* `3310` CA ALbers Equal Area

* `26910` UTM Zone 10, NAD83 (Northern Cal)

* `3857` Web Mercator (web maps)


## Transform all layers to UTM 10N, NAD83

Use `spTransform` to transform `sfhomes15_sp` and `bart` to `UTM 10N, NAD83`

- `highways` and `sfboundary` already have this CRS

Recall, this transformation is called `projecting` or `reprojecting`


## Transform all layers to UTM 10, NAD83

First, transform `sfhomes15_sp`

Note the two methods for doing same thing.

```{r}
sfhomes15_utm <- spTransform(sfhomes15_sp, CRS("+init=epsg:26910"))

# OR

# sfhomes15_utm <- spTransform(bart, CRS(proj4string(sfboundary)))

```

## BART data

The bart data is still a data.frame object.

So it needs to be 

1. to be converted to a SpatialPointsDataFrame
2. have its CRS defined
3. to be transformed to UTM CRS

Try it....

## BART data
```{r}

coordinates(bart) <- c("X","Y")

proj4string(bart) <-  CRS("+init=epsg:4326")

bart_utm <- spTransform(bart, CRS(proj4string(sfboundary)))
```


## Check

Do the CRSs all match?

```{r}
proj4string(bart_utm) 
proj4string(sfboundary)
proj4string(highways)
proj4string(sfhomes15_utm)
```

## Reproject the CRS AGAIN!

Use the CRS of an `existing sp object` if you want to get them to match exactly!
```{r}
#sfhomes15_utm <- spTransform(sfhomes15_sp, CRS("+init=epsg:26910"))

sfhomes15_utm <- spTransform(sfhomes15_sp, CRS(proj4string(sfboundary)))

# check
proj4string(sfhomes15_utm)
proj4string(sfboundary)
```


## Map all layers

Visual check

```{r}
plot(sfboundary)
lines(highways, col='purple', lwd=4)
points(sfhomes15_utm)
plot(bart_utm, col="red", pch=15, add=T)
```

# Spatial Analysis

## Spatial Analysis

1. Mapping / plotting to see location and distribution

2. Asking questions of, or querying, your data

3. **Cleaning & reshaping the data**

4. Applying analysis methods

5. Mapping analysis results

6. Repeat as needed

## Spatial Queries

There are two key types of spatial queries

- **spatial measurement** queries, 
    - e.g. area, length, distance


- **spatial relationship** queries, 
    - e.g. what locations in A are also in B.

These types are often combined, e.g.

- What is the area of region A is within region B?

# Spatial Measurement Queries

## `rgeos`

`rgeos` is THE package for manipulating geometric objects - or the geometry component of an `sp` object. 

We will use it to compute things like area and distance.

We will also use it to compare and intersect geometries and compute new geometries.

For detailed documentation see, `??rgeos`

## Computing Area

What is the area of San Francisco?

What data would we use to answer that question?

## Area of San Francisco

- Use `rgeos::gArea` to compute the area of SpatialPolygon objects

- What are the units?

- Check results against Wikipedia for [SF](https://en.wikipedia.org/wiki/San_Francisco)

```{r}
gArea(sfboundary)
```

## Area in sq km

[SF](https://en.wikipedia.org/wiki/San_Francisco)
 
```{r}
proj4string(sfboundary) # to get the CRS units

gArea(sfboundary) / (1000 * 1000) # Convert to square KM

```


## Length of highways

Use the `rgeos` function `gLength` to compute length of linear spatial objects
```{r}
gLength(highways) / 1000 # in kilometers
```

## Distance

The `rgeos` function `gDistance` will return the min distance between two geometries.

Compute the distance in kilometers between Embarcadero & Powell St Bart stations

- You can check on [Google Maps[(http://maps.google.com)] (always spot check!)
```{r}
gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
          bart_utm[bart_utm$STATION == 'POWELL STREET',]) /1000
```

## Distance between locations

Get distance between all `sfhomes` and `Embarcadero` BART

```{r}
dist2emb<- gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
                     sfhomes15_utm, byid=TRUE) /1000
# check output
nrow(dist2emb)
nrow(sfhomes15_utm)
head(dist2emb)
```

## Distance between locations

*How is the output different if `byid=FALSE`* (default)

?gDistance
```{r, eval=F}
#dist2emb<- gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
#                     sfhomes15_utm, byid=TRUE) /1000

gDistance(bart_utm[bart_utm$STATION == 'EMBARCADERO',],
                     sfhomes15_utm) /1000
```



# Spatial Relationships

## Spatial Relationships

**Spatial relationship queries** compare the geometries of two spatial objects in the same coordinate space (CRS).

There are many, often similar, functions to perform these queries (can be confusing!).

These operations may return logical values, lists, matrices, dataframes, geometries or spatial objects

- you need to check what type of object is returned 

- you need to check what values are returned to make sure they make sense



## Are there any BART stations in SF?

This is a very common type of spatial query called a `point-in-polygon` query.

We can use the `rgeos` function `gIntersects` to answer this.

We already know the answer but we want to see how it is done.


## Are there any BART stations in SF?

*What is the output of gIntersects?*

 - uncomment to find out
 
```{r, eval=FALSE}
bart_stations_in_sf <-gIntersects(bart_utm, sfboundary) 

# bart_stations_in_sf
```

## Are there any BART stations in SF?

What is the output of gIntersects?

- uncomment to find out!
```{r}
bart_stations_in_sf <-gIntersects(bart_utm, sfboundary)

bart_stations_in_sf
```

## WHAT Bart stations are in SF

Same function, but `byid=TRUE`
What is the output of gIntersects this time?

- uncomment to find out!

```{r}
sfbart_stations <-gIntersects(bart_utm, sfboundary, byid=TRUE)

# class(sfbart_stations)
# sfbart_stations
```


## What Bart stations are in SF

What is the output of gIntersects?

- why is it different?
```{r}
sfbart_stations <-gIntersects(bart_utm, sfboundary, byid=TRUE)

class(sfbart_stations)
sfbart_stations
```

*Why is it different?*

## What are the names of the SF BART stations?

We can use the result of `gIntersects` to view the names of the bart stations

```{r}
bart_utm[as.vector(sfbart_stations),]$STATION
```


## Subset the SPDF

We can use the result of `gIntersects` to subset `bart_utm`

```{r}
sfbart_utm <- bart_utm[as.vector(sfbart_stations),]
```



## Map the BART stations

```{r}
tmap_mode("view")

tm_shape(sfboundary) + 
  tm_polygons(col="beige", border.col="black") +
tm_shape(sfbart_utm) + 
  tm_dots(col="red")
```



## Reset `tmap` to plot mode


```{r}
tmap_mode("plot")
```

## gIntersects recap

`rgeos::gIntersects(geom1, geom2)` returns 

- returns a single TRUE/FALSE value indicating if any feature in geom1 intersects any feature in geom2 
    - if `byid=FALSE` (the default when not specified)


- returns a TRUE/FALSE matrix indicating what features in geom1 intersect features geom2 
    - if `byid=TRUE` 


## SF Census Tracts

Let's consider the `sfhomes15_utm` data along with SF Census tract data 


## SF Census Tracts

Use `readOGR` to load the shapefile `sftracts_wpop.shp`

- it's in the `data` folder

Explore the data

- What type of `sp` object is it?
- What is its CRS?
- What column contains the population data?

## SF Census Tracts
```{r}
sftracts <- readOGR(dsn="./data", layer="sftracts_wpop")

class(sftracts)
proj4string(sftracts)

```

## SF Census Tracts
```{r}
head(sftracts@data)
```

## Make a Quick Plot

```{r}
plot(sftracts)  
```

## Spatial Join

A spatial join associates rows of data in one object with rows in another object based on the spatial relationship between the two objects.

A spatial join is based on the comparison of two sets of geometries in the same coordinate space. 

 - This is also called a **spatial overlay**.

The default spatial relationship is `intersects`



## In what census tract is each property located?

We need to **spatially join** the `sftracts` and `sfhomes15_utm` to answer this.

What spatial object are we joining data from? to?

```{r, echo=F}
par(mfrow=c(1,2))
plot(sftracts)
plot(sfhomes15_utm)
```

 
## sp::over / rgeos::over

Spatial overlay operations in R are implemented using `over`

Simple point-in-polygon operations use `sp::over`

More complex geometric comparisons use `rgeos::over`

You need to have `rgeos` package installed!

## `over(x,y)`

`over(x,y)`

- for each feature in Layer X give me information about the first feature in Layer Y at the corresponding location. 

`over(x,y, returnList=TRUE)`  as:

- for each feature in Layer X give me information about the all features in Layer Y at the corresponding location. 

See `?over` for details.


## So here goes...

*In what census tract is each SF property located?* 

```{r, eval=F}
homes_with_tracts <- over(sfhomes15_utm, sftracts)
```

## Did it work?

If not, why not?


# Coordinate reference systems (CRS) must be the same!


## CRSs must be the same

The `over` function, like almost all spatial analysis functions, requires that both data sets be spatial objects (they are) with the same coordinate reference system (CRS). Let's investigate

```{r, eval=F}
# What is the CRS of the property data?
proj4string(sfhomes15_utm)

# What is the CRS of the census tracts?
proj4string(sftracts)
```


## Transform the CRS

```{r}
# Transform the CRS for tracts to be the same as that for sfhomes15_sp
sftracts_utm <- spTransform(sftracts, CRS(proj4string(sfhomes15_utm)))

# make sure the CRSs are the same
proj4string(sftracts_utm) == proj4string(sfhomes15_utm) 
```

Now let's try that overlay operation again

## Try 2

In what tract is each SF property is located?

- *Would a home be located in more than 1 tract?*
```{r}

homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)
```


## `over` output

What is our output? Does it answer our question?

What type of data object did the over function return?

Do we need to add `returnList = TRUE`

```{r, eval=F}
homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)

class(homes_with_tracts)
nrow(homes_with_tracts)
nrow(sftracts_utm)
nrow(sfhomes15_utm)
```

## Review the Over output

What do we have here?
```{r}
homes_with_tracts <- over(sfhomes15_utm, sftracts_utm)
class(homes_with_tracts)
nrow(homes_with_tracts)
nrow(sftracts_utm)
nrow(sfhomes15_utm)

```

## Review the over output
```{r}
head(homes_with_tracts)
```



## `over` discussion

Our output *homes_with_tracts* is a **data frame** that contains 

- the same number of rows as `sfhomes15_utm` - in same order
- the id of each property in `sfhomes15_utm`
- the data columns from `sftracts_utm@data` including the census tract id (GEOID) 

So we are close to answering our question.

But for the data to be useful we need 
- to add the `GEOID` column in `homes_with_tracts` to `sfhomes15_utm`

## Add the GEOID column

*CAUTION: this only works because the data are in the same order!*
```{r}
sfhomes15_utm$home_geoid <- homes_with_tracts$GEOID

# Review and note the change
#head(sfhomes15_utm@data)
```


## WOW

Data linkage via space!

The `over` operation gave us the census tract data info for each point in `sfhomes15_utm`

We added the `GEOID` for each point to the `@data` slot of `sfhomes15_utm`

We can now join `sfhomes15_utm` points by `GEOID` to any census variable, eg median household income, and then do an analysis of the relationship between, for example, property value and that variable.

**How would we do that?**

## Read in the census data

The `sf_med_hh_income2015.csv` file only has two columns: `GEOID` and `medhhinc`.

Because `GEOIDs` can have leading zeros, we set the `colClasses` to make sure they are not stripped.
```{r}
med_hh_inc <- read.csv("data/sf_med_hh_income2015.csv", stringsAsFactors = F, 
                       colClasses = c("character","numeric"))

head(med_hh_inc)
```

## Spatial*DataFrame to Data Frame Joins

We can use `sp::merge` to join the `med_hh_inc` DF to the `sfhomes15_utm` SPDF.

We should make sure that they share a column of common values - GEOID / home_geoid

We use `sp:merge` not regular `merge` to maintain the integrity of the `sp` object

  - i.e., make sure that the geometry and attribute data slots all stay aligned.

## Join by Attribute - a DF to a SPDF

Join two data objects based on common values in a column.

Use `sp:merge` to join a data frame to a spatial object with a data frames (`S*DF` objects)

```{r}

sfhomes15_utm <- merge(sfhomes15_utm, 
                       med_hh_inc, by.x="home_geoid", by.y="GEOID")
```

**IMPORTANT:** DO NOT merge a data frame to the @data slot! 

- Join it to the SPDF!

## Take a look at output

```{r}
head(sfhomes15_utm@data, 2)  # take a look with View(sfhomes15_utm@data)
```


## Check the `sp::merge` results

```{r}
tmap_mode("view")
tm_shape(sfhomes15_utm) + tm_dots(col="medhhinc")
```

## Reset plot mode
```{r}
tmap_mode("plot")
```

## The Census Tract Perspective

We now know the tract for each property.

Now let's think about this question from the tract perspective. 

Let's ask the question

- What is the average propety value per tract?



## Non-Spatial Aggregation

Since we joined GEOID to each property we can use the non-spatial `aggregate` function to compute the mean of totvalues for each GEOID.

But let's use the `sp` aggregate function instead!

It's actually more straight forward.

## sp::aggregate

Aggregate data values in one spatial object by the geometry of another using the specified function.

*What is the mean home value in each census tract?*
```{r}

tracts_with_mean_val <- aggregate(x = sfhomes15_utm["totvalue"], 
                                        by = sftracts_utm, FUN = mean)
```

Wow, so simple. What does that give us?


## Examine output of `sp::aggregate`

```{r}
class(tracts_with_mean_val)
head(tracts_with_mean_val@data)
nrow(tracts_with_mean_val) == nrow(sftracts_utm)
```

## sp::aggregate output

`sp::aggregate` returned a SpatialPolygonsDataFrame

The SPDF has the same geometry as `sftracts_utm`

But the `tracts_with_mean_val@data` slot only contains the mean totvalue for each tract.

To make these data more useful, let's add this value to `sftracts_utm`

## 

This only works because there are the same number of elements in both @data slots and they are in the same order!
```{r}

sftracts_utm$mean_totvalue <- tracts_with_mean_val$totvalue

head(sftracts_utm@data) # check it
```

## Map it 

Make a map of the results to make sure they seem reasonable.

First, set `tmap` to interactive mode
```{r}
tmap_mode("view")
```

## Map the results

This is called a 'choropleth' map.

```{r}
tm_shape(sftracts_utm) +
  tm_polygons(col="mean_totvalue", border.col=NA)
```

## Why no values for some tracts?
```{r}
tm_shape(sftracts_utm) +
  tm_polygons(col="mean_totvalue", border.col=NA) +
tm_shape(sfhomes15_utm) + tm_dots()
```

# Distance

## Distance queries

Many methods of spatial analysis are based on distance queries.

For example, point pattern analysis considers the distance between features to determine whether or not they are clustered.

We can also use distance as a way to select features spatially.

## Selecting by Distance

*What properties are within walking distance of BART?*

In order to select properties with 1KM of BART

- create a 1KM radius buffer polygon around each BART point

We then do a point-in-polygon operation to either count the number of properties within the buffer or compute the mean totvalue.

## rgeos

`rgeos` is the muscle for 

- creating new geometries from exisiting ones
- calculating spatial metrics like area, length, distance
- calculating the spatial relationship between two geometries.

We can use the `rgeos::gBuffer` function to create our buffer polygon

## Creating Buffers

The `rgeos::gBuffer` function takes as input a spatial object or objects to buffer and a buffer distance.

Let's assume 1KM is standard walking distance.
```{r}

bart_1km_buffer <- gBuffer(sfbart_utm, width=1000)

```

## Map the Buffer
```{r}
tmap_mode("view") 
tm_shape(bart_1km_buffer) + tm_polygons(col="red") +
tm_shape(sfbart_utm) + tm_dots()
```

## Try This

How does this differ when we add `byid=TRUE`
```{r}
bart_1km_buffer_byid <- gBuffer(sfbart_utm, width=1000, byid=TRUE)

tmap_mode("view")
tm_shape(bart_1km_buffer_byid) + tm_polygons(col="red") +
tm_shape(sfbart_utm) + tm_dots()


```

## What sfhomes are within 1km of any bart station?

What operation could we use?

Which buffer polygons?

##  What sfhomes are within 1km of any bart station?

Why `byid=TRUE`

What is the output?
```{r}
homes_near_bart <-  gIntersects(bart_1km_buffer, sfhomes15_utm, byid=TRUE) 
class(homes_near_bart)
head(homes_near_bart)
```

## Select the sfhomes15_utm points in the buffer

```{r}
# subset
sfhomes15_utm_near_bart <- sfhomes15_utm[as.vector(homes_near_bart),]
```

## Map it
```{r}
tm_shape(bart_1km_buffer) + tm_polygons(col="red") +
tm_shape(sfhomes15_utm_near_bart) + tm_dots()
```


## Questions


## Summary

That was a whirlwind tour of just some of the methods of spatial analysis.

There was a lot we didn't and can't cover.

`sf` package is emerging and will eclipse `sp`

Raster data is a another major topic!
- but the `raster` package is the key


## Selected  References & Tutorials

Introductory Tutorials

- [Spatial Data in R tutorial](https://cengel.github.io/rspatial)
- [NEON Spatial Data tutorials](http://neondataskills.org/tutorial-series/)
- [GIS in R](http://www.nickeubank.com/gis-in-r)

Geodata Visualization emphasis

- [Tmap in a Nutshell](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html)
- [Intro to visualizing Spatial Data in R](https://github.com/Robinlovelace/Creating-maps-in-R)
- [RStudio Leaflet in R tutorial](https://rstudio.github.io/leaflet)
- [Blog on mapping census data in R](http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/)

## Selected  References & Tutorials

Deep dive Tutorials that include spatial analysis

- [An Introduction to Spatial Data Analysis and Visualisation in R](https://data.cdrc.ac.uk/tutorial/an-introduction-to-spatial-data-analysis-and-visualisation-in-r)
- [Intro to GIS and Spatial Analysis (see appendices)](https://mgimond.github.io/Spatial/index.html)
- [Spatial Data Analysis and Modeling in R](http://www.rspatial.org/index.html)
- [Geocomputation in R (featuring sf Package)](http://robinlovelace.net/geocompr/ )
- [Weight Spatial Polygon Overlay tutorial, aka areal interpolation](http://rstudio-pubs-static.s3.amazonaws.com/6577_3b66f8d8f4984fb2807e91224defa854.html)


CRAN Spatial Packages

- [CRAN Task View: Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html)

## Output code to script

```{r, eval=F}
library(knitr)
purl("r-geospatial-workshop-pt2.Rmd", output = "scripts/r-geospatial-workshop-pt2-May-2018.R", documentation = 2)
```


