---
title: "R Geospatial Fundamentals-Raster: Appendix"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R Spatial Libraries

Let's load the libraries we will use

```{r, message=F, warning=F}
library(sf)     # simple features objects and methods
library(tmap)   # mapping spatial objects
library(raster) # reading in and operating on rasters
library(here)   # build paths from folder with .rproj file
```

# Appendix - Extra Exercise

> Note: this material may need updating.

## Predicting home values

Not at the census-tract level. But that's kind of coarse...

What if we want to do the analysis at the property level?

We have added this workflow as an appendix at the end of this document. It puts together a bunch of what we've learned over Parts I to III of the workshop. Feel free to give it try later!

```{r}
# Take a random subset of our 2015 homes, to save on computer time
SFhomes15_sample <- SFhomes15_sf[sample(seq(nrow(SFhomes15_sf)),
                                 replace = FALSE, size = 2000), ]
```

## Predicting home values

Now let's reproject all our data to a UTM projection, so that we units of meters rather 
than decimal degrees.

```{r}
# reproject
SFhomes15_utm <- st_transform(SFhomes15_sample, 26910)
DEM_utm <- projectRaster(DEM,
                         projectExtent(DEM,
                                       st_crs(SFhomes15_utm)$proj4string))
nlcd_green_utm <- projectRaster(nlcd_green,
                                projectExtent(nlcd_green,
                                              st_crs(SFhomes15_utm)$proj4string))

#check projections
st_crs(SFhomes15_utm) == st_crs(DEM_utm)
st_crs(SFhomes15_utm) == st_crs(nlcd_green_utm)
```

## Predicting home values

Now let's buffer all our homes with a 100-meter buffer, then sum the greenspace within
those buffers.

```{r}
#create buffer
SFhomes15_utm_buff <- st_buffer(SFhomes15_utm, dist = 100)

#sum the greenspace within the buffers 
#NOTE: This will take a couple minutes to run...
greenspace_homes <- extract(nlcd_green_utm, SFhomes15_utm_buff, 
                            fun = mean, 
                            na.rm = T)

#add that as a column in our SFhomes15_utm dataframe
SFhomes15_utm$greenspace = greenspace_homes
```

## Predicting home values

And now let's extract the elevation at each home.

```{r}
#extract the elevation to the homes
#NOTE: no need for fun or na.rm arguments here, because the homes
#and points, not polygons, so only a single cell will extract to each
elev_homes <- extract(DEM_utm, SFhomes15_utm)

#add that as a column in our SFhomes15_utm dataframe too
SFhomes15_utm$elev = elev_homes

```

## Predicting home values

**Now**, how about that regression model?

```{r}
mod <- lm(totvalue ~ elev + greenspace, data = SFhomes15_utm)
```

## Predicting home values

Mehhh... Some signal for elevation, but a very low $R^{2}$.

But again, this is **just a stand-in analysis** (and lacks various considerations and adjustments for statistical validity).

```{r}
summary(mod)
```

## Fog data

Data on summertime coastal fog in CA (a.k.a. Karl)
```{r}
# (from http://climate.calcommons.org/dataset/monthly-summertime-fog)
# (units are in average hours per day)
karl_files <- unique(gsub('.aux.xml', 
                          '', 
                          list.files(here("notebook_data",
                                          "Part_3",
                                          "CalMnYr"))))

karl_files <- karl_files[grep('flcc', karl_files)]

# take  a look
karl_files
```

```{r}
r1 <- raster(here("notebook_data",
                          "part_3","USGS_1_n38w123_20210615.tiff"))
```

## RasterStack

Read all of the KARL files into one RasterStack object
```{r}
karl <- stack(paste0(here("notebook_data",
                          "part_3",
                          "CalMnYr"), 
                     karl_files,))

karl <- stack(paste0('./notebook_data/part_3/data/CalMnYr/', karl_files))

# look at what we made!
karl
```
**A RasterStack object** -  literally what it sounds like!

## Plot the Stack

```{r}
#plot a few
plot(karl)
```

## CRS

What's the projection of the RasterStack?
```{r}
# what's the projection?
st_crs(karl)
```

## CRS Transformation

Let's reproject this
```{r}
karl_WGS = projectRaster(karl, projectExtent(karl, st_crs(SFtracts)$proj4string))

# check resultant CRS
st_crs(karl_WGS) == st_crs(SFtracts)
```


##  What did we get?

What type of spatial object did that give us?

```{r}
karl_WGS
```

## RasterBrick

A RasterBrick. What does that mean?
```{r, eval = FALSE}

# See the documentation!
?raster::brick

```


## Crop the Brick

Crop it to the extent of our area of interest - SF
```{r}
# Crop it to SFtracts
karl_WGS_crop = crop(karl_WGS, SFtracts)

#Note that R vectorized that operation across our entire RasterBrick, the same way that it vectorizes many operations, e.g. 3 < 4 vs 3 < seq(4).
```


## Plot it

```{r}
# now let's make our same plot again
par(mfrow = c(1,2))
plot(karl_WGS[[7]])
plot(st_geometry(SFtracts), add = T, reset=F, key.pos = NULL, col = NA)
plot(karl_WGS_crop[[7]])
plot(st_geometry(SFtracts), add = T, reset = F, key.pos = NULL, col = NA)
```

## Mean Fog

Let's mean the karl values across the RasterBrick
```{r}
# Mean values
mean_karl_WGS_crop <- mean(karl_WGS_crop)
```

What did that give us?
```{r}
mean_karl_WGS_crop
```

## RasterBrick to Raster

When we computed the mean of a RasterBrick we got back a RasterLayer object! That makes sense, because we took cellwise means across all Layers in our Brick.

**This is called raster algebra**

## Plot it

```{r}
plot(mean_karl_WGS_crop)
plot(st_geometry(SFtracts), add = T, col = NA)
```


## BUT

Not all common operations successfully run as raster algebra
```{r, eval=FALSE}
# This won't work
sd_karl_WGS_crop = sd(karl_WGS_crop)
```

Let's try this instead
```{r}
sd_karl_WGS_crop <- calc(karl_WGS_crop, sd)
```

## Plot it

```{r}
# plot that too
par(mfrow = c(1,2))
plot(mean_karl_WGS_crop, main = "mean summer fog hours")
plot(st_geometry(SFtracts), add = T, reset=F, key.pos=NULL, col = NA)
plot(sd_karl_WGS_crop, main = "sd summer fog hours")
plot(st_geometry(SFtracts), add = T, reset=F, key.pos=NULL, col = NA)
```

## Thoughts...

Looks like the foggiest neighborhoods also have the highest variation in fog, 
but some less foggy neighborhoods on the east side of the city
(e.g. The Inner Mission, Bayview) also vary quite a bit

## Extract Mean Values

Extract fog values to our tracts
```{r}
SFtracts$mean_karl <- extract(mean_karl_WGS_crop, 
                              SFtracts, 
                              mean)
```


## Explore Spatial Relationships

Maybe fogginess is a function of elevation? That would make sense, right?

```{r}
# Linear regression model
mod = lm(mean_karl ~ mean_elev, data = SFtracts)
```

## View results
```{r}
summary(mod)
```

## Comments

Neat!

Census-tract mean elevation has a significant, positive effect on
mean fogginess (with each meter increase in elevation causing on average
a .011 hour, or about 39-second increase in time spent in fog on an average
summer day.

Though this barely explains about 3.5% of the variance in our data!

Of course, again, these are __not statistically valid models!__ Just analytical stand-ins.
