---
title: "Geospatial Data & Maps in R, pt 3"
author: "Patty Frontiera and Drew Hart, UC Berkeley D-Lab"
date: "November 2018"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
editor_options: 
  chunk_output_type: console
---




---------------------------
# TODO:


1. add explanation of some packages not converting to , and demo of how to convert objects to  objects, in Day 3
           world_sp = as(world, Class = "Spatial")
                   world_sf = st_as_sf(world_sp)

2. decide whether or not to produce and use the tutorial files


----------------------------





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part III Prep

1. https://github.com/dlab-geo/r-geospatial-workshop

- Click *Clone or Download* and download the zip file
- Upzip the zip file and make a note of the folder in which it is located

2. Open RStudio and start a **new script** or continue the one from last week

3. Set your working directory to where you downloaded and unzipped the files

4. Follow along by opening `r-geospatial-workshop-pt3.html` in a web browser

5. Make sure required libraries are installed. 

- sp, rgdal, rgeos, tmap, raster

## Follow Along

Open one of the tutorial files in a web browser

Slides
`r-geospatial-workshop-pt3.html `

R Code
`scripts/r-geospatial-workshop-pt3.Rmd`

*Make sure you can cut and paste into RStudio*

## Part III Overview

Review basic raster concepts

Read in our previous spatial data and new raster data

Make some raster and combined raster-vector maps

Run some raster and raster-vector operations and analyses

## R Spatial Libraries

Let's load the libraries we will use

```{r, eval=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read/write from file; manage CRSs
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
library(raster) # reading in and operating on rasters
```

##

```{r, echo=FALSE}
library(sp)     # spatial objects and methods
library(rgdal)  # read and write from file
library(rgeos)  # geometric operations
library(tmap)   # mapping spatial objects
library(raster) # reading in and operating on rasters
```
 
## Set your working directory

Use `setwd` to set your working directory to the location of the tutorial files.

For example:

```{r, eval=FALSE}
path = "/path/to/your/working_directory"
setwd(path)
```

# Load some files from Parts I and II

## SF Census Tracts

```{r}
# Read in the 'sftracts_wpop' shapefile
tracts = readOGR('./data', 'sftracts_wpop')
```

## SF Properties 2015

```{r}
# Read in from CSV file
print(list.files())
sfhomes <- read.csv('./data/sf_properties.csv', 
                    stringsAsFactors = FALSE)
# subset the data
sfhomes15 <- subset(sfhomes, as.numeric(SalesYear) == 2015)

sfhomes15_sp <- sfhomes15  # Make a copy

# Make it spatial
coordinates(sfhomes15_sp) <- c('lon','lat')

#Assign it a proj4string using the EPSG code
proj4string(sfhomes15_sp) <- CRS("+init=epsg:4326")

#Reproject to the tracts projection
#NOTE: We're overwriting the previous sfhomes15_sp object here! This is
#fine to do if we want, but we should always beware.
sfhomes15_sp = spTransform(sfhomes15_sp, CRS(proj4string(tracts)))

#check projection equality
proj4string(sfhomes15_sp) == proj4string(tracts)

```

# SECTION I: Raster Data in R

## Read in a Raster

Using the fabulous `raster` package

```{r}
# Reading in and plotting raster files

#read in a Bay Area DEM (Digital Elevation Model)
#(from http://www.webgis.com/terr_pages/CA/dem1/sanfrancisco.html)
DEM = raster('./data/san_francisco-e.DEM')
```

## Plot it

Make a quick map of the data to check it out
```{r}
#plot it
plot(DEM)
```

## Explore the Structure

To see a summary of the structure and content of a `RasterLayer` object,
we can just enter the object's name.

```{r}
DEM
```
You can view more detail in the RStudio Environment pane.

## Explore the Structure

Note that this display is formatted very similarly to the display for
a `Spatial*DataFrame` object. There's a good amount of consistency across
the main geospatial packages in R.

```{r}
tracts
```

## Explore the Structure

A raster should really just be a matrix of data, with metadata assiging
it the correct location on the Earth's surface, right?
## Explore the Structure

Well, here's some of the important metadata.

```{r}
DEM@extent
DEM@crs
DEM@ncols
```

## Explore the Structure

Note that this is very similar to the metadata of a 
`Spatial*DataFrame` object (although we get `@extent`
and `@crs` slots for what are called `@bbox` and 
`@proj4string` in a `Spatial*DataFrame` object).

```{r}
tracts@bbox
tracts@proj4string
```

## Explore the Structure

Also note that, because of the distinct nature of raster and vector data,
`RasterLayer` and `Spatial*DataFrame` objects have some fundamentally different metadata.

For example, a `RasterLayer` has information on the resolution.

```{r}
DEM
```

## Explore the Structure

What does the resolution mean?

```{r}
(DEM@extent@xmax - DEM@extent@xmin) / DEM@ncols
(DEM@extent@ymax - DEM@extent@ymin) / DEM@nrows
DEM
```
It's the cell-size! That is, the real-world distance of the x and y sides of
each cell (i.e. pixel) in our `RasterLayer`, expressed in the distance units
of the object (as determined by its projection).

## Explore the Structure

As for our matrix of data, it should live somewhere inside the `@data` slot, right?
(Just like our data.frame of attribute data lives in the `Spatial*DataFrame@data` slot.)

```{r}
DEM@data
```

## Explore the Structure

There's a bunch of stuff there. Let's take a look at the `@data@values` slot...

```{r}
DEM@data@values
```
## Explore the Structure

What happened there? How can our data be `logical(0)`?
Where's our data?

```{r}
DEM@data@inmemory
DEM@data@fromdisk
```

It appears our data was not all read into memory. Instead, our `RasterLayer` object
contains a pointer to where that data is held on disk (to save memory).

## Explore the Structure

How do we see the values, then?

Well, we can still subset the values, using two-dimensional subsetting notation, just
as we'd expect for a matrix.

```{r}
DEM[10:15, 20:30]
```
And we'll get back a vector of the values from that subsetted matrix, unfolded.

## Explore the Structure

We can even subset the entire thing, getting back all of the values.

```{r}
DEM[,]
```

## Explore the Structure

And we could even then turn that vector back into a matrix of the proper dimensions.

```{r}
#coerce our whole raser's dataset to a matrix, with the appropriate number
#of columns, and 
matrix(DEM[,], ncol = ncol(DEM), byrow = TRUE)
```

Notice that we were able to use `ncol` on our `RasterLayer` object, because in some 
basic senses it behaves like a `matrix`, much as a `Spatial*DataFrame` behaves in some
senses like a `data.frame`.

## Explore the Structure

Can we create a new `RasterLayer` object from a matrix?

```{r}
test = raster(matrix(DEM[,], ncol = ncol(DEM), byrow = TRUE))
test
```

We do get a new `RasterLayer`! However, notice that it of course has no
projection, and has incorrect extent and resolution.

## Explore the Structure

You can imagine that you could then use
the DEM object's information to assign the new `RasterLayer` its appropriate metadata
(much like what we did with unprojected point data that we read in from a CSV and
coerced to a `SpatialPointsDataFrame`). 

Because there's a much easier way to do all of this...

## Explore the Structure

Our subsetting operation can actually take a `drop = FALSE` argument!

```{r}
DEM[10:15, 20:30, drop = FALSE]
```

And this gives us back a new RasterLayer object! 

In essence, this makes the subsetting operation more or less like a basic **clipping** function.

## Explore the Structure

Here's what that gave us:

```{r}
test = DEM[10:15, 20:30, drop = FALSE]
plot(test)
```

## Explore the Structure

And now what does the `@data@values` slot look like in our new object now?

```{r}
test@data@values
```

## Explore the Structure

That makes sense, because our test object didn't come from a file, so R
didn't leave its data on disk.

```{r}
test@data@inmemory
test@data@fromdisk
```

## CRS Transformations

Check the CRS (aka projection) of the DEM raster data.

Transform (or reproject) the Tracts vector data to match it.

```{r}
#check out its projection
proj4string(DEM)

#reproject tracts to our DEM projection
tracts_NAD = spTransform(tracts, CRS(proj4string(DEM)))
```

## CRS Transformations

You can reproject the DEM with `raster::projectRaster`

Note how the syntax is similar to `sp::spTransform`

```{r}
DEM_WGS = projectRaster(DEM, projectExtent(DEM, CRS(proj4string(tracts))))

# Now let's check equivalence
proj4string(tracts_NAD) == proj4string(DEM)
proj4string(DEM_WGS) == proj4string(tracts)
```

## Clipping Rasters

Since the raster data covers a larger area than our vector data / area of interest, we can clip it.

```{r}
# clip the WGS CRS version of the rasters to tracts 
DEM_WGS_crop = crop(DEM_WGS, extent(tracts))

# Clip the NAD CRS version
DEM_crop = crop(DEM, extent(tracts_NAD))
```

## Plot Clipped Raster data

You can plot raster and vector
```{r}
plot(DEM_WGS_crop)
 
```

## Plot Raster & Vector Data

And we can plot raster and vector data together!
Here we plot the NAD CRS version
```{r}
#plot together
plot(DEM_WGS_crop)
plot(tracts, add = T)
```

## Masking Raster Data

Notice that the clipping (or cropping) operation reduced our dataset
to just the extent of the census-tract dataset, but still left us with
values outside of the census tracts themselves (because they are areas outside
the city of San Francisco).

## Masking Raster Data

For some purposes, we may want to get rid of those
values as well. 

We can do this with an operation called **masking**.

```{r}
DEM_WGS_crop_masked = mask(DEM_WGS_crop, tracts)
```

## Masking Raster Data

Here's what that gives us, compared to the unmasked object:

```{r}
DEM_WGS_crop_masked
DEM_WGS_crop
```

Notice that this is still a rectangular grid of cells, with the same `nrow` and `ncol`
as the unmasked object. This is, of course, because a raster will **always** be a
rectangular grid of cells. What masking did was to set the cells that lie outside
our dataset to NAs.

## Masking Raster Data

```{r}
plot(DEM_WGS_crop_masked)
plot(tracts, add = T)
```

## Plotting rasters in TMAP

We can make an interactive plot using `TMAP`

```{r}
my_map <- tm_shape(DEM_WGS_crop_masked) +
  tm_raster() +
tm_shape(tracts) + 
  tm_borders() +

# Set mode to interactive
tmap_mode("view")
```

## View Map
```{r}
my_map
```

## Writing Raster Data

And now that we've manipulated our data as desired, we can write it to disk
if we like!

```{r}
#write our reprojected, cropped data to the data directory, using the Geotiff format
#(and allow R to overwrite if file already exists)
writeRaster(DEM_WGS_crop_masked, filename="./data/DEM_reproject_crop.tif", format="GTiff", overwrite = T)
```



# SECTION II: Raster Operations and Spatial Analysis

## Challenge section

This section will feature a number of challenges, to get you practicing some of the
material we've already covered today.

## Extract elevation values

We can use the raster::extract function to get the elevation values for each tract.

```{r}
# get the elevation for every cell in each of the census tracts
elev = extract(DEM_WGS_crop, tracts)

#what did that give us?
head(elev)

```


## What is the output?

Check out the data in the `elev` object
```{r}
length(elev)
nrow(tracts)
```

## What is the output?

A vector of the elevations for all the cells within each census tract!

```{r}
length(elev)
nrow(tracts)
```

## Average Raster value by Vector shape

Let's get each tract's average elevation
```{r}
mean_elev = lapply(elev, mean, na.rm = T)
head(mean_elev)

```

## Add elevation to Vector object

Let's add this to the tracts_NAD df
NOTE: the order remains the same, so we can just add this right in!
```{r}
tracts$mean_elev = unlist(mean_elev)
```


## Map it
```{r}
#what did we get?
elev_map <- tm_shape(tracts) + 
  tm_polygons(col = 'mean_elev') +
  tm_layout("The pain of biking in SF, by census tract", 
            inner.margins=c(0,0,.1,0), title.size=4.8)
```

## Map it

```{r}
elev_map
```

## One step Re-do

We can also pass a function argument to `raster::extract`
```{r}
elev = extract(DEM_WGS_crop, tracts, fun=mean)

#what did that give us?
head(elev)
```

## Questions?

Can you see the difference between mapping a vector layer on top of a raster layer (my_map) compared to summarizing raster values by vector polygons and then mapping the polygon features (elev_map)?


## Challenge 1: Read in and check out new data

You have another raster dataset in your `./data` directory. The file is called
`nlcd2011_sf.tif`. This is data from the National Land Cover Database (NLCD; https://www.mrlc.gov/nlcd11_leg.php). It's 2011 data that was downloaded from https://viewer.nationalmap.gov/basic.
Read that file in as an object called `nlcd`, and plot it.

## Work in progress...

Reminder: file is called `nlcd2011_sf.tif`

## Solution

```{r}
#read in nlcd data
nlcd = raster('./data/nlcd2011_sf.tif')

#plot nlcd
plot(nlcd)
```

## Let's see what's in the NLCD data

```{r}
freq(nlcd)
```

## ... and a barplot

(Please note that the colors in this barplot have no relations to the colors used to plot our raster! Pay attention
instead to the numbered categories along the x-axis, instead.)

```{r}
barplot(nlcd)
```

## What do those values mean?

This is a categorical raster. Each cell on the raster holds a discrete (integer) value, coding a particular type of land-cover (rather than a continuous value, like we saw with our elevation data above).

Where do we do to figure out what the codes mean? This should come with the metadata that ships with your data. [Here] (https://www.mrlc.gov/nlcd11_leg.php) is a layout of the NLCD codes.


## Challenge 2: Reproject and crop our NLCD data

Now that we've read in our NCLD data, check if we need to reproject it (we want it to be in the same projection as our `tracts` object), and project it if need be. Then crop it to the extent of our `tracts` object.

## Work in progress...

## Solution

```{r}
#check projection equality
proj4string(nlcd) == proj4string(tracts)

#reproject
nlcd_WGS = projectRaster(nlcd, projectExtent(nlcd, CRS(proj4string(tracts))))

#check projection equality again
proj4string(nlcd_WGS) == proj4string(tracts)

#crop
nlcd_WGS_crop = crop(nlcd_WGS, extent(tracts))

```

## Plot the new raster

```{r}
plot(nlcd_WGS_crop)
```

## Recovering our plot formatting

Notice that the colors of our original, which conveniently represented the NLCD class
colors from their website, are lost after reprojecting and cropping our raster.

Those colors were actually all controlled by a bunch of information stored in the original
NLCD file, which was read into the `@legend` slot of our RasterLayer object when we read that file in. Here's the information -- and note that the codes that look like '#435ae2', for example, are hexadecimal strings, a common way of representing colors in the programming world:

```{r}
nlcd@legend
```

## Recover our plot formatting

What's the `@legend` slot in our reproject, cropped object look like?

```{r}
nlcd_WGS_crop@legend
```

## Recover our plot formatting

Well that's a bummer...

Is there a way we could transfer that information over to our reprojected, cropped raster?

## Challenge 3: Transferring our `@legend` info
Try to figure out how to transfer the `@legend` info from the original raster object to our new object. Then plot the new object to see if it worked.

## Work in progress...

## Solution

```{r}
nlcd_WGS_crop@legend = nlcd@legend
plot(nlcd_WGS_crop)

```
Nice! Not necessary... but nice!


## Reclassifying rasters

When we're working with a categorical raster, we'll often want to reclass our data. We may want to do this because:

- Our original data has more classifications than we actually need for our analysis.
- We want to represent the classifications we do have by a different numerical scheme because it somehow makes our analysis more convenient.

## Reclass the NLCD

Let's reclass our NLCD data. First we'll need to define a reclassification
matrix with 3 columns (low, high, to):

```{r, eval = FALSE}
?reclassify
```

## Define reclassification matrix 
```{r}
reclass_df <- c(0, 12, NA, # water will be set to NA (i.e. 'left out' of our analysis)
                20, 21, 1, # we'll treat developed open space as greenspace, based on NLCD description
                21, 30, 0, # developed and hardscape will have 0s
                30, 32, NA, 
                40, Inf, 1) # greensapce will have 1s
reclass_df
reclass_m <- matrix(reclass_df, ncol = 3, byrow = TRUE)
reclass_m
```

## Reclassify the raster 
```{r}
nlcd_green <- reclassify(nlcd, reclass_m)
```

## Reclassify the raster

What did we get?

```{r}
freq(nlcd_green)
```

## Reclassify the raster 

What did we get?

```{r}
barplot(nlcd_green)
```

## Reclassify the raster 

What did we get?

```{r}
plot(nlcd_green)
```

## Challenge 4: Extract our NLCD data to our tract polygons

Just like we did earlier with our elevation data, let's extract our reclassed
NLCD data to our census-tract polygons.

## Work in progress...

## Solution

```{r}

#extract the mean nlcd_simple values to tract polygons
greenspace = c(extract(nlcd_green, tracts, fun=mean))

```

## What did we get?

```{r}
greenspace
```
## Why?

What's with all the NAs?

Remember, we set all water cells to NA, to ignore them in our analysis.

## Challenge 5: How do we get extract to ignore our NAs?

Try to run the same command again, but telling the extract function to ignore NAs.

## Work in progress...

## Solution

The `na.rm` argument will do this for us. See how the docs indicate that it is set to FALSE by default? 

```{r, eval = FALSE}
?raster::extract
```

Good to know!  Also good to know that this is a common
argument across a variety of R operations.

## Solution

```{r}
#extract the mean nlcd_simple values to tract polygons,
#this time setting na.rm to TRUE
greenspace = extract(nlcd_green, tracts, fun=mean, na.rm = T)

#and add to our tracts dataframe (which we can do because order is preserved)
tracts$prop_greenspace = greenspace

```
## Get the mean home values in each tract

Pulling code from the end of Part II of this workshop, let's aggregate our homes data to the tract level too.

```{r}
#aggregate totvalue to tracts
tracts_w_mean_val = aggregate(x = sfhomes15_sp['totvalue'], by = tracts, FUN = mean)
```

## Get the mean home values in each tract

```{r}
#use a quick tmap to check that it looks like sensible output
qtm(tracts_w_mean_val, fill = 'totvalue')

#and add the mean_val column to our tracts dataframe
tracts$mean_totvalue = tracts_w_mean_val$totvalue

```

## Predicting home values

Do mean elevation and proportion greenspace predict mean total home values in each tract?

```{r}
mod = lm(mean_totvalue ~ mean_elev + prop_greenspace, data = tracts)
summary(mod)
```

## Predicting home values

Not at the census-tract level. But that's kind of coarse...

What if we want to do the analysis at the property level?

Here's a workflow that puts together a bunch of what we've learned over
Parts I to III of the workshop!

```{r}
#First, we'll take a random subset of our 2015 homes, so that our analysis doesn't take so long to compute.
sfhomes15_sample = sfhomes15_sp[sample(seq(nrow(sfhomes15_sp)), replace = FALSE, size = 2000), ]
```

## Predicting home values

Now let's reproject all our data to a UTM projection, so that we units of meters rather 
than decimal degrees.

```{r}
#reproject
sfhomes15_utm <- spTransform(sfhomes15_sample, CRS("+init=epsg:26910"))
DEM_utm = projectRaster(DEM, projectExtent(DEM, CRS(proj4string(sfhomes15_utm))))
nlcd_green_utm = projectRaster(nlcd_green, projectExtent(nlcd_green, CRS(proj4string(sfhomes15_utm))))

#check that the projections are all good
proj4string(sfhomes15_utm) == proj4string(DEM_utm)
proj4string(sfhomes15_utm) == proj4string(nlcd_green_utm)

```

## Predicting home values

Now let's buffer all our homes with a 100-meter buffer, then sum the greenspace within
those buffers.

```{r}
#create buffer
sfhomes15_utm_buff = gBuffer(sfhomes15_utm, width = 100, byid = T)

#sum the greenspace within the buffers 
#NOTE: This will take a couple minutes to run...
greenspace_homes = extract(nlcd_green_utm, sfhomes15_utm_buff, fun = mean, na.rm = T)

#add that as a column in our sfhomes15_utm dataframe
sfhomes15_utm$greenspace = greenspace_homes
```

## Predicting home values

And now let's extract the elevation at each home.

```{r}
#extract the elevation to the homes
#NOTE: no need for fun or na.rm arguments here, because the homes
#and points, not polygons, so only a single cell will extract to each
elev_homes = extract(DEM_utm, sfhomes15_utm)

#add that as a column in our sfhomes15_utm dataframe too
sfhomes15_utm$elev = elev_homes

```

## Predicting home values

**Now**, how about that regression model?

```{r}
mod = lm(totvalue ~ elev + greenspace, data = sfhomes15_utm)
summary(mod)

```

Mehhh... Some signal for elevation, but very low R^2.

Hopefully it's a demonstrative analysis though!


## Questions?




# SECTION III: RasterStacks and RasterBricks

## RasterStacks

What about working with multiple rasters?

## Fog data

Data on summertime coastal fog in CA (a.k.a. Karl)
```{r}
#(from http://climate.calcommons.org/dataset/monthly-summertime-fog)
#(units are in average hours per day)
karl_files = unique(gsub('.aux.xml', '', list.files('./data/CalMnYr')))
karl_files = karl_files[grep('flcc', karl_files)]

# Take  a look
karl_files
```

## RasterStack

Read all of the KARL files into one RasterStack object
```{r}
karl <- stack(paste0('./data/CalMnYr/', karl_files))

# look at what we made
karl

```
*A RasterStack object -  literally what it sounds like!*

## Plot the Stack

```{r}
#plot one
plot(karl[[7]])
plot(tracts, add = T)
```

## CRS

What's the projection of the RasterStack?

## CRS

What's the projection of the RasterStack?
```{r}

#what's the projection?
proj4string(karl)
```

## CRS Transformation

Let's reproject this
```{r}
karl_WGS = projectRaster(karl, projectExtent(karl, CRS(proj4string(tracts))))

# check resultant CRS
proj4string(karl_WGS) == proj4string(tracts)

```


##  What??

What type of spatial object did that give us?

```{r}
karl_WGS
```

## RasterBrick

A RasterBrick. What does that mean?
```{r, eval = FALSE}

# See the documentation!
?raster::brick

```


## Crop the Brick

Crop it to the extent of our area of interest - SF
```{r}
# Crop it to tracts
karl_WGS_crop = crop(karl_WGS, extent(tracts))

#Note that R vectorized that operation across our entire RasterBrick, the same way that it vectorizes many operations, e.g. 3<4 vs 3< seq(4)
```


## Plot it

```{r}
# now let's make our same plot again
par(mfrow = c(1,2))
plot(karl[[7]])
plot(tracts, add = T)
plot(karl_WGS_crop[[7]])
plot(tracts, add = T)


```

## Mean Fog

Let's mean the karl values across the RasterBrick
```{r}
# Mean values
mean_karl_WGS_crop = mean(karl_WGS_crop)
```

What did that give us?
```{r}
mean_karl_WGS_crop
```

## RasterBrick to Raster

When we computed the mean of a RasterBrick we got back a RasterLayer object! That makes sense, because we took cellwise means across all Layers in our Brick.

**This is called raster algebra**

## Plot it

```{r}
plot(mean_karl_WGS_crop)
plot(tracts, add = T)
```


## BUT

Not all common operations successfully run as raster algebra
```{r, eval=FALSE}
# This won't work
sd_karl_WGS_crop = sd(karl_WGS_crop)
```

Let's try this instead
```{r}
sd_karl_WGS_crop = calc(karl_WGS_crop, sd)
```

## Plot it

```{r}
#plot that too
par(mfrow = c(1,2))
plot(mean_karl_WGS_crop)
plot(tracts, add = T)
plot(sd_karl_WGS_crop)
plot(tracts, add = T)
```

## Thoughts...

Looks like the foggiest neighborhoods also have the highest variation in fog, 
but some less foggy neighborhoods on the east side of the city
(e.g. The Inner Mission, Bayview) also vary quite a bit

## Extract Mean Values

Extract fog values to our tracts
```{r}
tracts$mean_karl = extract(mean_karl_WGS_crop, tracts, mean)
```


## Explore Spatial Relationships

Maybe fogginess is a function of elevation? That would make sense, right?

```{r}
# Linear regression model
mod = lm(mean_karl ~ mean_elev, data = tracts)
```

## View results
```{r}
summary(mod)
```

## Comments

Neat! Census-tract mean elevation has a significant, positive effect on
mean fogginess (with each meter increase in elevation causing on average
a .011 hour, or about 39-second increase in time spent in fog on an average
summer day. Though this only explains about 4% of the variance in our data!

## Questions?

**Additional Resources**
 https://cran.r-project.org/web/packages/raster/vignettes/Raster.pdf 

 https://www.neonscience.org/raster-data-r

 https://nceas.github.io/oss-lessons/spatial-data-gis-law/4-tues-spatial-analysis-in-r.html

 https://geoscripting-wur.github.io/IntroToRaster/
